Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities

### 2021-10-15 paper/manuscript.Rmd
Minor changes
Response to Ben comments

```{r echo=FALSE}
read_csv(here::here("figures/man_model_table.csv")) %>%
  select(metric, everything()) %>%
  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric),
         group = case_when(
           group == "rarefyID:study_id" ~ "timeseries:study",
           group == "study_id" ~ "study",
           group == "Residual" ~ "residual",
           TRUE ~ group),
         term = case_when(
           term == "(Intercept)" ~ "Intercept",
           term == "year_scaled" ~ "Year",
           term == "sd__(Intercept)" ~ "Intercept sd",
           term == "cor__(Intercept).year_scaled" ~ "Intercept Year corr",
           term == "sd__year_scaled" ~ "Year sd",
           term == "sd__Observation" ~ "Observation sd",
           TRUE ~ term
         )) %>%
  mutate(across(c(estimate, std.error, statistic, df, p.value), round, 2), 
         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
  arrange(metric) %>%
  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness (S) and standardized metrics. Additional model estimates can be found in the supplement.") %>%
  collapse_rows(columns = 1, valign = "middle") %>%
  kable_classic() %>% 
  kable_styling(latex_options= c("scale_down", "hold_position"))
```

@@

While we did not directly measure changes in rare species, our results also contradict likely scenarios of loss predicted due to rare species extinction.
Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species cannot be replaced by functionally similar species [@leitão2016] .
Instead, we observed patterns consistent with species replacement by functionally redundant species from the species pool.

What does local maintenance of functional structure mean for ecosystem function?
The vast majority of experimental and observational work links declines in function to declines in functional or species diversity [@brose2016; @cadotte2011; @duffy2007].
By those criteria very few communities in our dataset are in a state of concern for loss of functionality.
However, the dimensions of functional space most important for ecosystem function are still a topic of ongoing debate, and at least some known aspects important for multifunctionality were not measured here [e.g. dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
The fact that we observed so many communities maintaining structure across the most commonly used metrics for linking biodiversity to function calls into question how previous work translates to natural communities.
Metrics need to be both closely linked to changes in ecosystem function and also experiencing shifts in natural communities to be meaningful.

@@

Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
However, a number of gaps in best practices may be obscuring a true underlying trend.
First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
Most time series span only a few years (Figure 1) and may not provide the statistical power necessary to detect trends.
The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias. We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021]. We do however emphasize that even disturbed communities maintain functional structure [@edwards2013; @matuoka2020], and the inclusion of more heavily impacted sites would not necessarily change the result.

@@

Our results illustrate a well known but increasingly pervasive problem in synthesis-based ecology of distinguishing true ecological patterns from methodological limitations.
We present here the conclusion best supported by available data with a number of caveats and suggestions for next steps forward.
The most pressing need is for intentional and targeted data collection efforts.
We join others in the call for increased monitoring in under sampled areas and continued efforts to centralize existing data sources [@gonzalez2016; @hughes2021; @vellend2017].
Data that fill geographic, taxonomic and trait gaps should be prioritized over further collection of data that replicate existing biases.
One relatively low-cost high-reward data investment is collation of additional species-level trait means.
Intentional trait selection is critical for linking functional patterns to ecological processes [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases.
Traits would be especially beneficial if they explicitly incorporate existing frameworks for linking traits to processes, including the now widely accepted response and effect framework [@lavorel2002].

<!-- We also need theory that links ecological process to the kinds of changes actually happening in communities. Do the shifts we see matter? Are there shifts we didn't even measure that are critical for function and resilience? -->

>>>>

```{r echo=FALSE}
read_csv(here::here("figures/man_model_table.csv")) %>%
  select(metric, everything()) %>%
  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric),
         group = case_when(
           group == "rarefyID:study_id" ~ "timeseries:study",
           group == "study_id" ~ "study",
           group == "Residual" ~ "residual",
           TRUE ~ group),
         term = case_when(
           term == "(Intercept)" ~ "Intercept",
           term == "year_scaled" ~ "Year",
           term == "sd__(Intercept)" ~ "Intercept sd",
           term == "cor__(Intercept).year_scaled" ~ "Intercept Year corr",
           term == "sd__year_scaled" ~ "Year sd",
           term == "sd__Observation" ~ "Observation sd",
           TRUE ~ term
         )) %>%
  mutate(across(c(estimate, std.error, p.value), round, 2), 
         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
  arrange(metric) %>%
  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness (S) and standardized metrics. Additional model estimates can be found in the supplement.") %>%
  collapse_rows(columns = 1, valign = "middle") %>%
  kable_classic() %>% 
  kable_styling(latex_options= c("scale_down", "hold_position"))
```

@@

While we did not directly measure changes in rare species, our results also contradict likely scenarios of loss predicted due to rare species extinction.
Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020]. Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species can eventually no longer be replaced by functionally similar species [@leitão2016].
While we did not have a large enough time window in most time series to capture the complete extinction process, observed patterns were instead consistent with species replacement by functionally redundant species from the species pool.

What does local maintenance of functional structure mean for ecosystem function?
The vast majority of experimental and observational work links declines in function to declines in functional or species diversity [@brose2016; @cadotte2011; @duffy2007]. By those criteria very few communities in our dataset are in a state of concern for loss of functionality. However, shifts in metrics are only relevant if the underlying traits are those most critical for ecosystem function. We were limited in this analysis to the traits available rather than those with strong empirical links to function. Similarly, the dimensions of functional space most important for ecosystem function are still a topic of ongoing debate, and at least some known aspects important for multifunctionality were not measured here [e.g. dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
Still, the fact that we observed so many communities maintaining structure across the most commonly used metrics for linking biodiversity and function calls into question how previous work translates to natural communities.
Metrics need to be both closely linked to changes in ecosystem function and also experiencing shifts in natural communities to be meaningful.

@@

Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
However, a number of gaps in best practices may be obscuring a true underlying trend.
First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
Most time series span only a few years (Figure 1) and may not provide the statistical power necessary to detect trends.
The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias. We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021], particularly scenarios of conversion to urban, human-dominated landscapes. While evidence from other work shows even disturbed communities can maintain functional structure [@edwards2013; @matuoka2020], these results should not be interpreted as evidence of low functional impact in areas of heavy human disturbance.

@@

Here we make a significant first step in establishing a general trend for functional diversity through time across a variety of taxa and systems.
We present the conclusion best supported by available data and acknowledge that it is still too early to confidently distinguish between true ecological pattern and methodological limitations.
The most pressing next step is for intentional and targeted data collection efforts.
We join others in the call for increased monitoring in under sampled areas and continued efforts to centralize existing data sources [@gonzalez2016; @hughes2021; @vellend2017]. Data that fill geographic, taxonomic and trait gaps should be prioritized over further collection of data that replicate existing biases. One relatively low-cost high-reward data investment is collation of additional species-level trait means. Intentional trait selection is critical for linking functional patterns to ecological processes [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases. Trait collection should explicitly consider existing frameworks for linking traits to processes [e.g. the response and effect framework @lavorel2002] to facillitate clear ecological interpretation of potential functional changes.
